---
title: "[DITK] Local Outlier Factor"
output: html_notebook
---

# Pre-liminaries

### Installing the packages needed.

Please update and run here when adding new packages

```{r}
pkgs <- sort(c('readr', 'dplyr', 'tidyr', 'tidylog', 'lubridate', 'ggplot2', 'gridExtra', 'cowplot', 'ggmap', 'leaflet', 'viridis', 'DataCombine', 'ggplotify', 'ggpubr', 'tinytex', 'Rlof', 'plotly', 'cluster'))

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)){
  install.packages(pkgs_install)
}
library(readr)
library(dplyr)
library(tidyr)
library(tidylog)
library(lubridate)
library(ggplot2)
library(gridExtra)
library(cowplot)
library(ggmap)
library(leaflet)
library(viridis)
library(htmltools)
library(DataCombine)
library(ggplotify)
library(ggpubr)
library(tinytex)
library(Rlof)
library(plotly)
library(cluster)
```

# Data Preprocessing Continuation

### Importing the datasets

We import the dataset using `read_csv` and use `view` `head` and `tail` to view some records.

```{r}
merged_df <- read_csv("datasets/preprocessed.csv")
```

```{r}
df2 <- another <- merged_df
```

### Optional: Rounding to 2 decimal places

```{r}
df2 <- round(df2[, c(2:9)], digits = 2)
df2$DATE <- another$DATE
df2 <- df2[, c(9, 1:(8))]
```

### Optional: Scaling the Data

```{r}
numdataset <- subset(df2, select = -c(DATE))
scaled_dataset <- as.data.frame(scale(numdataset))
```

### Optional: Splitting the Data

```{r}
df2 <- another <- merged_df
df2 <- round(df2[, 2:9], digits = 2)
df2$DATE <- another$DATE
last_col <- names(df2)[ncol(df2)]
df2 <- cbind(df2[, last_col], df2[,-ncol(df2)])
```

```{r}
df2$YEAR <- year(df2$DATE)
df2$MONTH <- month(df2$DATE)
df2$DAY <- day(df2$DATE)
df2 <- subset(df2, select = -c(DATE))
```

```{r}
head(df2)
```

```{r}
yearly_data <- split(df2, df2$YEAR)
head(yearly_data$`2023`)
```

```{r}
View(yearly_data$`2023`)
```

# Local Outlier Factor

```{r}

lof_scores <- Rlof::lof(scaled_dataset[, c(3, 5)], k = 10)
threshold <- 1.4
 
# Identify and mark outliers
outliers <- scaled_dataset[lof_scores > threshold, ]
df2$OUTLOF <- ifelse(lof_scores > threshold, "Outlier", "Inlier")
df2$LOF <- lof_scores
 
 
lof1 <- ggplot(df2, aes(`AVE. HUM`, `AVE. TEMP`, color = OUTLOF)) +
  geom_point() +
  scale_color_manual(values = c("skyblue", "lightcoral")) +
  geom_text(aes(`AVE. HUM`, `AVE. TEMP`, label = ifelse(OUTLOF == "Outlier", paste(`AVE. TEMP`, "," , `AVE. HUM`), "")),  # Add labels only for outliers
            vjust = 1,  # Adjust vertical position of text (optional)
            hjust = 1,  # Adjust horizontal position of text (optional)
            inherit.aes = FALSE) +  # Inherit aesthetics except color
  theme_minimal() +
  labs(title = "Weather Anomaly Detection with LOF")

lof1
```

```{r}
fig <- plot_ly(df2, x = ~DATE, y = ~`AVE. TEMP`, z = ~`AVE. HUM`, color = ~OUTLIER, colors = c('skyblue', 'lightcoral'), marker = list(size = 5))
fig <- fig %>% add_markers()
fig <- fig %>% layout(scene = list(xaxis = list(title = 'Date'),
           yaxis = list(title = 'Average Temperature'),
           zaxis = list(title = 'Average Humidity')))
fig
```

```{r}
df2[df2$OUTLIER == "Outlier", ]

```

```{r}
# Assuming lof_scores is the vector containing LOF scores

# Histogram plot
histogram <- ggplot(scaled_dataset, aes(lof_scores)) +
  geom_histogram(bins = 30, color = "steelblue") +  # Adjust bins as needed
  labs(title = "Distribution of LOF Scores", x = "LOF Score", y = "Frequency") +
  theme_minimal()

# Density plot
density_plot <- ggplot(scaled_dataset, aes(lof_scores)) +
  geom_density(color = "royalblue") +
  labs(title = "Density of LOF Scores", x = "LOF Score", y = "Density") +
  theme_minimal()


grid.arrange(histogram, density_plot, nrow = 1)
```

```{r}
ggsave("plots/lof1.png", plot = lof1, dpi = 300, width = 10, height = 6)
ggsave("plots/densityplot.png", plot = density_plot, dpi = 300, width = 10, height = 6)
ggsave("plots/histogram.png", plot = histogram, dpi = 300, width = 10, height = 6)
```

# Evaluating

### Visual Inspection

```{r}
VI <- df2
```

```{r}
plot1 <- plot_ly(
  x = VI$DATE,  # Replace 'date' with your time variable
  y = VI$`AVE. TEMP`,
  type = 'scatter',
  mode = 'lines',
  name = 'Average Temperature in time series'
)

plot1 <- add_lines(plot1,
                    x = VI$DATE,
                    y = VI$`AVE. HUM`,
                    type = 'scatter',
                    mode = 'lines',
                    name = 'Average Humidity',
                   )

plot1 <- add_trace(plot1,
                    x = VI$DATE,
                    y = VI$LOF,
                    type = 'scatter',
                    mode = 'markers',
                    name = 'LOF Scores',
                    color = c("skyblue", "lightcoral", "lightgreen")[VI$OUTLOF == "Inlier" ]
                   )

plot1
```

### Silhouette Analysis

Calculate distance matrix

```{r}
dist_matrix <- as.matrix(dist(scaled_dataset[, c(3, 5)]))
```

```{r}
sil <- silhouette(df2$LABELS, dist_matrix)
```

```{r}
plot(sil)
```

### Using Distance Metrics

```{r}
dm <- df2
```

```{r}
dist_matrix <- as.matrix(dist(scaled_dataset[, c(3, 5)]), method = "euclidian")
```

```{r}
# Calculate the mean distance to k-nearest neighbors for each point
dm$KNN <- apply(dist_matrix, 1, function(row) {
    sorted_distances <- sort(row)  # Sort distances for each point
    mean(sorted_distances[2:(10)])  # Calculate mean distance to k-nearest neighbors
})
```

```{r}
# Set a threshold to determine outliers based on mean distance to k-NN
threshold <- 0.29
 
# Identify and mark outliers
dm$OUTKNN <- ifelse(dm$KNN > threshold, "Outlier", "Inlier")
# Adding the new column OUTBOTH based on the conditions
dm$OUTBOTH <- ifelse(dm$OUTKNN == "Inlier" & dm$OUTLOF == "Inlier", 1, # Condition for 1
              ifelse(dm$OUTLOF == "Outlier" & dm$OUTKNN == "Inlier", 2, # Condition for 2
              ifelse(dm$OUTKNN == "Outlier" & dm$OUTLOF == "Inlier", 3, # Condition for 3
              ifelse(dm$OUTLOF == "Outlier" & dm$OUTKNN == "Outlier", 4, NA)))) # Condition for 4

```

```{r}
# Plot LOF scores against mean distance to k-nearest neighbors
plot(k_nearest_neighbors, lof_scores, xlab = "Mean Distance to k-Nearest Neighbors", ylab = "LOF Score",
     main = "LOF Scores vs Mean Distance to k-Nearest Neighbors")

```

```{r}
plot <- plot_ly(dm, 
                x = ~KNN, y = ~LOF, 
                type = 'scatter', mode = 'markers', color = ~OUTBOTH, colors = c('skyblue', 'lightcoral', 'lightgreen', 'mediumpurple')) %>%
    layout(title = "LOF Scores vs Mean Distance to k-Nearest Neighbors",
           xaxis = list(title = "Mean Distance to k-Nearest Neighbors"),
           yaxis = list(title = "LOF Score"))

plot
```

```{r}
plot <- plot_ly(dm, 
                x = ~`AVE. HUM`, y = ~`AVE. TEMP`, 
                type = 'scatter', mode = 'markers', color = ~OUTBOTH, colors = c('skyblue', 'lightcoral', 'lightgreen', 'mediumpurple')) %>%
    layout(title = "LOF Scores vs Mean Distance to k-Nearest Neighbors",
           xaxis = list(title = "Mean Distance to k-Nearest Neighbors"),
           yaxis = list(title = "LOF Score"))

plot
```
